---
title: "Fuzzy design RDD"
title-block-banner: false
subtitle: "Causal statistics for treatment models"
author:
  - name: Olivier CARON
    email: caron.olivier.80@gmail.com
    affiliations: 
      name: "Paris Dauphine-PSL"
      city: Paris
      state: France
date: "last-modified"
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
  echo: false
---


```{r}
#| label: load-data

library(arrow) # For reading parquet files
library(ggplot2)
library(dplyr)
library(ggstatsplot) # For ggcoefstats
library(ivreg) # For 2SLS
library(ggthemr)
library(huxtable) # for lovely tables
library(rdrobust) # For robust RDD estimation (sharp and fuzzy)

ggthemr("greyscale")
data <- arrow::read_parquet("data/EduloanFEMALE.parquet")

theme_custom <- theme(
  text = element_text(size = 12),
  panel.grid = element_blank(),
  panel.grid.major.y = element_line(
    colour = "#e3e1e1",
    linetype = 2
  ),
  plot.title.position = 'plot',
  legend.position = 'top',
  legend.title = element_blank()
)
```

## Data description

```{r}
#| label: data-summary
#| echo: false
#| eval: true

skimr::skim(data)

```
\newpage

## Research Question

::: {.callout-note title="Research Question"}
We want to know if receiving a loan (the treatment T) increases the chances to enroll at university (the outcome Y), using an RDD.
:::

## Fuzzy RDD Methodology

We implement a Fuzzy Regression Discontinuity Design (RDD) due to the following institutional setup:

- The Empirica score is used as the eligibility criterion for student loans.
- In theory, students with `Empirica > 0` should receive a loan.
- In practice, this rule is not strictly applied—some students below the threshold get loans, and some above do not—creating a "fuzzy" discontinuity.

Thus, we treat the threshold crossing as an instrument for actual loan receipt, using it in a Two-Stage Least Squares (2SLS) setup.

## Visual Evidence

We begin by visually inspecting the discontinuity at the threshold.


```{r}
#| label: data-plot7
#| layout-ncol: 2

# Filter the data to the optimal bandwidth [-13, 13] given by the professor
data_optimal <- data %>% 
  filter(empirica >= -13 & empirica <= 13)

# Adding binned averages to the first stage plot (with uniform point sizes)
bin_data <- data_optimal %>%
  mutate(bin = cut(empirica, breaks = seq(-13, 13, by = 1))) %>%
  group_by(bin) %>%
  summarize(
    mean_empirica = mean(empirica),
    mean_loan = mean(loan),
    mean_univ_register = mean(univ_register)
  )

# Proper RDD visualization
# First-stage visualization (treatment probability)
p1 <- ggplot(data, aes(x = empirica, y = loan)) +
  geom_smooth(data = subset(data, empirica < 0), method = "lm", se = TRUE) +
  geom_smooth(data = subset(data, empirica >= 0), method = "lm", se = TRUE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "First Stage: Loan Approval Probability",
       x = "Empirica Score",
       y = "Probability of Receiving Loan") +
  ylim(0, 1) +
  xlim(-13, 13) +
  theme_custom +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1))

p1 <- p1 + 
  geom_point(data = bin_data, aes(x = mean_empirica, y = mean_loan), 
             color = "red", size = 1, alpha = 0.2)

# Reduced form visualization (outcome)
p2 <- ggplot(data, aes(x = empirica, y = univ_register)) +
  geom_smooth(data = subset(data, empirica < 0), method = "lm", se = TRUE) +
  geom_smooth(data = subset(data, empirica >= 0), method = "lm", se = TRUE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Reduced Form: University Enrollment",
       x = "Empirica Score",
       y = "Probability of University Enrollment") +
  ylim(0, 1) +
  xlim(-13, 13) +
  theme_custom +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1))

p2 <- p2 + 
  geom_point(data = bin_data, aes(x = mean_empirica, y = mean_univ_register), 
             color = "red", size = 1, alpha = 0.2)
print(p1)
print(p2)


```

The figure above illustrates the discontinuity at the Empirica score cutoff (0) for both loan approval and university enrollment:

- Left panel: There is a clear and sharp increase in the probability of receiving a loan just above the threshold. This confirms that the Empirica score functions effectively as an instrument, even if imperfectly enforced.

- Right panel: A visible jump in university enrollment at the same cutoff mirrors the pattern observed in the first stage, suggesting a potential causal link between loan receipt and enrollment decisions.


## Regressions
### First Stage Regression + Reduced Form Regression + 2SLS Estimation

```{r}
#| label: regression-first-stage

# First Stage Regression (using local linear regression)
first_stage <- lm(loan ~ discont_r + empirica + empirica_r, data = data_optimal)
```

```{r}
#| label: regression-reduced-form

# 2 Reduced Form Regression
reduced_form <- lm(univ_register ~ discont_r + empirica + empirica_r, data = data_optimal)
```
\newpage
```{r}
#| label: iv-regression

# 4. 2SLS Estimation
iv_model <- ivreg(univ_register ~ loan + empirica + empirica_r | discont_r + empirica + empirica_r, 
                  data = data_optimal)

# Summary with huxreg, it's better than stargazer, flextable, etc.

huxreg(
  "FS (Loan ~ Threshold)" = first_stage,
  "RF (Univ ~ Threshold)" = reduced_form,
  "2SLS (Univ ~ Loan)" = iv_model,
  statistics = c("N" = "nobs", "R²" = "r.squared") # we can also add "LogLik" = "logLik", "AIC" = "AIC"
)
```









### Key Regression Results and Interpretation

| Model                       | Key Coefficient       | Estimate | Interpretation                                                                 |
|----------------------------|------------------------|----------|---------------------------------------------------------------------------------|
| First Stage Regression     | `discont_r`            | 0.422    | Students just above the Empirica threshold (repayment score > 0) are 42.2% more likely to receive a loan, confirming the instrument's strength. |
| Reduced Form Regression    | `discont_r`            | 0.202    | Students just above the Empirica threshold are 20.2% more likely to enroll in university, suggesting a positive reduced-form effect of loan eligibility. |
| 2SLS Estimation (IV model) | `loan`                 | 0.478    | Receiving a loan increases the probability of university enrollment by 47.8%. This reflects the causal effect for compliers identified by the fuzzy RDD. |

\newpage
## Confirm results with Robust Estimation (rdrobust)

```{r}
#| label: rdrobust-fuzzy-plots
#| layout-ncol: 2
#| echo: false

# First Stage plot with automatic bin selection (IMSE-optimal, quantile-spaced)
test1 <- rdplot(
  y = data$loan, 
  x = data$empirica,
  binselect = "qs", # best visualization with quantile-spaced bins
  x.lim = c(-13, 13), 
  y.lim = c(0, 1),
  title = "First Stage (Loan) - Optimal Bins",
  x.label = "Empirica Score",
  y.label = "Probability of Receiving Loan",
  masspoints = "off"
)

# Reduced Form plot with automatic bin selection
test2 <- rdplot(
  y = data$univ_register, 
  x = data$empirica,
  binselect = "qs",
  x.lim = c(-13, 13), 
  y.lim = c(0, 1),
  title = "Reduced Form (University Registration) - Optimal Bins",
  x.label = "Empirica Score",
  y.label = "Probability of University Registration",
  masspoints = "off"
)
```

Let's now estimate the effect of the loan on university registration using the rdrobust package.

```{r}
#| label: rdrobust-fuzzy

rd_result_fuzzy <- rdrobust(y = data$univ_register,
                            x = data$empirica,
                            fuzzy = data$loan,  # indique que c'est une fuzzy RDD
                            c = 0,
                            p = 1, q = 2,
                            kernel = "tri")

summary(rd_result_fuzzy)


#bw <- rdbwselect(y = data$univ_register,
                 #x = data$empirica,
                 #c = 0,  # cutoff
                 #p = 1,  # local linear
                 #q = 2,  # bias-correction polynomial
                 #kernel = "tri", 
                 #bwselect = "mserd")
#print(bw)

# Pour les bandes passantes h (gauche et droite)
#h_values <- bw$bws[1, 1:2]

# Pour les bands passantes b (gauche et droite)
#b_values <- bw$bws[1, 3:4]

#rd_result <- rdrobust(y = data$univ_register,
                      #x = data$empirica,
                      #c = 0,
                      #p = 1, q = 2,
                      #h = h_values,
                      #b = b_values,
                      #kernel = "tri")

#rd_result

#summary(rd_result)
```
\newpage

| Method      | Estimate | Statistically Significant? | Method comparison                                                                    |
|-------------|----------|-----------------------------|------------------------------------------------------------------------------------|
| `rdrobust`  | 0.468    | Yes (p = 0.001)             | LATE estimated using robust local polynomial regression at the Empirica threshold[^1]. |
| 2SLS        | 0.478    | Yes (p = 0.0024)            | LATE estimated using a global linear Two-Stage Least Squares (2SLS) specification.         |

Both methods yield consistent estimates of approximately 47%.  

[^1]: When using `rdrobust` for a fuzzy RDD, we need to specify the argument `fuzzy = data$loan`. Otherwise, the package will treat the RDD as a sharp design.

## Conclusion

Fuzzy Regression Discontinuity Design analysis reveals a substantial causal effect of student loans on university enrollment among South African female students: obtaining a loan increases the likelihood of university enrollment by 47%.






```{r}
#| label: data-plot6
#| fig-cap: "Fuzzy RDD"
#| echo: false
#| eval: false

## Bonus 
ggcoefstats(
  x = iv_model,
  title = "2SLS Regression: Effect of Loan on University Registration"
)

```


```{r}
#| label: data-plot2
#| fig-cap: "Effect of Loan on University Registration"
#| echo: false
#| eval: false

ggbarstats(
  data = data,
  x = loan,
  y = univ_register,
  title = "",
  xlab = "Loan",
  ylab = "University Registration"
)

#This bar chart shows the link between getting a loan and enrolling in university.

#- Among students **without a loan**, only **33%** registered.
#- Among those **with a loan**, **57%** registered.
 
#This difference is highly significant (**$\chi^2(1) = 269.33$, $p < 0.001$**) with a moderate effect size (**Cramér’s $V = 0.23$** $95\%$ CI: $[0.20, 0.26]$).
#Each group also shows significant differences in registration rates compared to what we could expect by chance ($p < 10^{-15}$), supporting the idea that loans help students enroll.

```


